Up to 25\% of beneficial mutations in protein sequences are not adaptive innovations in mammals

### 2022-11-04 manuscript/main.tex
Reorganized the manuscript, substitutions then SNPs

\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\textrm{Var}}
\newcommand{\der}{\textrm{d}}
\newcommand{\e}{\textrm{e}}
\newcommand{\avg}[1]{\left< #1 \right>} % for average
\newcommand{\Ne}{N_{\textrm{e}}}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\textrm{fix}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\Sphy}{S}
\newcommand{\Sphyclass}{\mathcal{C}}
\newcommand{\SphyMean}{\overline{\Sphy}}
\newcommand{\divStrongDel}{\Sphy < -3}
\newcommand{\divDel}{-3 < \Sphy < -1}
\newcommand{\divWeakDel}{-1 < \Sphy < 0}
\newcommand{\divWeakAdv}{0 < \Sphy < 1}
\newcommand{\divAdv}{ \Sphy > 1}
\newcommand{\PdivStrongDel}{\proba \left[ \divStrongDel \right]}
\newcommand{\PdivDel}{\proba \left[ \divDel \right]}
\newcommand{\PdivWeakDel}{\proba \left[ \divWeakDel \right]}
\newcommand{\PdivWeakAdv}{\proba \left[ \divWeakAdv \right]}
\newcommand{\PdivAdv}{\proba \left[ \divAdv \right]}
\newcommand{\given}{\mid}
\newcommand{\Spop}{\beta}
\newcommand{\SpopMean}{\overline{\Spop}}
\newcommand{\polyDel}{\Spop < -1}
\newcommand{\polyNeutral}{-1 < \Spop < 1}
\newcommand{\polyAdv}{ \Spop > 1}
\newcommand{\PpolyDel}{\proba \left[ \polyDel \right]}
\newcommand{\PpolyNeutral}{\proba \left[ \polyNeutral \right]}
\newcommand{\PpolyAdv}{\proba \left[ \polyAdv \right]}

@@

\title{In mammals, an important proportion of advantageous mutations are repairing damaged genomes instead of creating adaptive innovations}

@@

    \begin{abstract}
        Positive selection can be the result of adaptive evolution, driven by a change in environment or selective pressure.
        However, positive selection can also be the result of what we call maintenance positive selection where advantageous mutations are reverting deleterious substitutions without any change of the environment.
        The evolutionary outcomes of these two positive selections are drastically different.
        While adaptive evolution acts as a diversifying force, maintenance positive selection stabilizes phenotypes and reduces diversity across species.
        By leveraging polymorphism data in 28 populations and divergence data between 94 mammalian species, we estimate that around 25\% of observed substitutions are owed to maintenance positive selection.
        Our work confirms that deleterious substitutions have accumulated in mammals and are currently being eliminated, resulting in widespread positive selection that is not adaptive evolution.
    \end{abstract}

    \keywords{Selection \and phylogenetic \and population genetics \and codon models \and adaptation }

    %\NS{on devrait plus placer le départ de l'intro dans un contexte historique, pour dire que de connaître si une mutation est avantageuse est la base de la génétique moléculaire (en citant des papiers clés)}
    Adaptation is considered as the main process driving the diversity of forms and functions in the tree of life\cite{darwin_origin_1859}.
    It allows species to live in new ecological niches, and to evolve in response to a change in environment\cite{darwin_origin_1859, merrell_adaptive_1994}.
    The notion of adaptation, in its eco-evolutionary sense, is tightly linked to the notion of change, both of the environment (biotic and abiotic), and of the species responding to the latter\cite{merrell_adaptive_1994}.
    With the discovery of DNA, the main support of heritable information, evolutionary geneticists looked for signatures of adaptation in genomes, such as to quantify it\cite{mcdonald_adaptative_1991, eyre-walker_distribution_2006}.
    In that context arose the term of rate of adaptive evolution\cite{mcdonald_adaptative_1991, eyre-walker_distribution_2006, galtier_adaptive_2016}.
    However, in contrast to adaptation in its eco-evolutionary sense, adaptive evolution is merely used as a synonymous of evolution under positive selection\cite{eyre-walker_distribution_2006}.
    This leads to a misunderstanding on what is adaptation with the loss of the notion of change, and a misinterpretation of the so-called rate of adaptive evolution.
    Indeed, adaptive evolution is not the only process which can lead to positive selection at the population scale\cite{sella_application_2005, mustonen_fitness_2009}.
    At mutation-selection-drift equilibrium, a mutation restoring the ancestral allele which corresponded to an optimal fitness in a constant environment, but lost due to the fixation of a deleterious mutation by drift, will also be positively selected (fig.~\ref{fig:fitness-landscape}B).
    In this study, this process will be referred to as maintenance positive selection.
    %At the scale of a population, adaptive evolution and maintenance positive selection display different dynamics.
    %While in the case of environmental change, the population as a whole is climbing the fitness landscape along an adaptive walk\cite{tenaillon_utility_2014}, in the case of maintenance positive selection, the population as a whole is at mutation-selection equilibrium and therefore merely stays at the same location in the fitness landscape\cite{sella_application_2005, mustonen_fitness_2009}.
    %We can distinguish two main scenarios that can lead to such positive selection.
    %In the first scenario, a mutation towards a state which was deleterious is now advantageous because of a change of environment.
    %A change of environment could be a change of temperature, viral exposition but it could also be a change in another part of the species' genome.
    %In the new environment, the mutation allows its bearer to survive and reproduce more than the others and thus will be positively selected.
    %This process is typically referred to as adaptive evolution (fig.~\ref{fig:fitness-landscape}A), and meets the eco-evolutionary definition of adaptation.
    At the population scale, maintenance positive selection and adaptive evolution are indistinguishable and both result in a positive transmission bias of the advantageous allele, but their consequences at a macro-evolutionary scale are fundamentally different.
    While adaptive evolution promotes phenotype diversification (fig.~\ref{fig:fitness-landscape}C), maintenance positive selection promotes phenotype stability and preserves well established biological systems (fig.~\ref{fig:fitness-landscape}D).
    It is thus essential to distinguish these two types of positive selection to understand the long term effects of selection on the diversification of species.

    At the gene level, positive selection can be quantified by comparing sequence variation inside a population with the sequence divergence from a sister species\cite{mcdonald_adaptative_1991}.
    The basic idea is that advantageous mutations are expected to be observed in substitutions between species but not in polymorphism within a population because they reach fixation too quickly.
    This method has been largely improved to account for possible bias-inducing factors such as the presence of weak positive selection, background selection and demographic fluctuations\cite{eyre-walker_distribution_2006, eyre-walker_estimating_2009, galtier_adaptive_2016, tataru_inference_2017}.
    It has been applied across many organisms and enabled the discovery of widespread positive selection\cite{moutinho_variation_2019}.
    However, the exact causes for this widespread occurrence of positive selection have been scarcely investigated\cite{eyre-walker_distribution_2006, eyre-walker_estimating_2009, galtier_adaptive_2016} %\NS{is it the case in the conclusion of the previous paper?? would be nice to back it up with references}.
    As mentioned above, not all advantageous mutations contribute to genetic innovation and diversification, and truly adaptive evolution can only be well estimated if maintenance positive selection is taken into account.
    While adaptive evolution is unpredictable because it caused by an unpredictable change in the environment of organisms, the second is predictable because optimal states should be more or less the same for closely related organisms which share most of their environment and physiology.
    A way to thus measure maintenance positive selection is to predict selection coefficients assuming a constant and stable fitness landscape estimated from the alignment of protein-coding genes.
    For each site of the protein alignment, a first approximation would be to consider that the frequency of an amino acid is a proxy for its fitness, since for example a deleterious amino acid will be more likely to be absent in the alignment.
    One of the most used methods to make these predictions is SIFT (Sort Intolerant From Tolerant), which reliably detects deleterious mutations in many taxa\cite{ng_sift_2003, vaser_sift_2016}.
    However, this method has two major shortcomings.
    First, we know that a lot of mechanisms can influence amino acid frequency other than fitness (phylogeny, mutation biases, GC-biased gene conversion, mutational paths) which are not accounted for in SIFT\@.
    Second, SIFT scores is not interpretable as a selection coefficient.
    Contrarily, so-called mutation-selection codon models are rooted within a population genetics formalism\cite{halpern_evolutionary_1998, mccandlish_modeling_2014}.
    These phylogenetic codon models can disentangle processes of mutation at the DNA level and selection at the amino acid level.
    Such codon models provide a nearly-neutral framework of evolution by estimating the fitness landscape over amino acid sequences for each site of the sequence (fig~\ref{fig:method}A) while taking into account non-adaptive processes impacting codon frequency\cite{halpern_evolutionary_1998, rodrigue_mechanistic_2010, tamuri_estimating_2012, latrille_improved_2022}.
    From the fitness landscape obtained by fitting the model to the data, one can thus compute the scaled selection coefficient ($\Sphy$) of any non-synonymous mutations.
    Importantly, mutation-selection codon models assume that the underlying fitness landscape is stable along the phylogeny (as in fig.~\ref{fig:fitness-landscape}B), thus a mutation with a positive $\Sphy$ would be driving the current site toward a fitter amino-acid, and such mutation is effectively considered under maintenance positive selection.

    By integrating datasets at the species and population level, we can predict $\Sphy$ for each mutation currently segregating in a population (fig~\ref{fig:method}B).
    Such a dataset can allow us to address different questions regarding positive selection affecting protein-coding DNA sequences.
    Which proportion of the segregating polymorphisms in population are predicted toward a more optimal amino acid (positive $\Sphy$), thus supposedly under maintenance positive selection?
    Are these polymorphisms effectively under positive selection at the scale of populations?
    Finally, how much of positive selection observed in populations is due to maintenance positive selection?
    To answer these questions, we estimated the fitness landscape of amino-acids for long-term evolution using available multiple sequence alignments between mammalian species\cite{ranwez_orthomam_2007, howe_ensembl_2021}.
    We then integrated data at the phylogenetic and population scale to estimate divergence and polymorphism, respectively, across the entire exome for 28 populations from 6 genera (\textit{Equus},  \textit{Bos}, \textit{Capra}, \textit{Ovis}, \textit{Chlorocebus} and \textit{Homo}; fig.~\ref{fig:method}A\&B).

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.fitness-landscape}
        \caption{
            Panel A and B: at a given site inside a protein-coding DNA sequence, the different amino acids (x-axis) have different fitnesses (y-axis).
            Under a fluctuating fitness landscape (panel A), these fitnesses are changing with time.
            The protein sequence is always lagging behind the moving target defined by the amino acid fitnesses, and since substitutions are accepted preferentially if they are in the direction of this target, substitutions are on average adaptive.
            At the phylogenetic scale (panel C), this phenomena promotes phenotype diversification across species.
            Under a stable fitness landscape (panel B), the protein sequence only fluctuates at the mutation-selection balance.
            All the mutations reaching fixation are either slightly deleterious and reaching fixation due to drift or are advantageous and restoring a more optimal amino acid.
            At the phylogenetic scale (panel D), this phenomena promotes promotes phenotype stability and preserves well established biological systems.
        }
        \label{fig:fitness-landscape}
    \end{figure*}

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.method}
        \caption{
            Integrating divergence and polymorphism.
            At the phylogenetic level (panel A), the amino acid Wrightian fitness for each site is estimated from protein-coding DNA alignments using mutation-selection codon models.
            At the population-genetic level (panel B), for each observed single nucleotide polymorphism (SNP) segregating in the population, selection coefficient ($\Sphy$) are computed as the difference in amino acid fitness between the ancestral and derived variants.
            Mutation-selection codon models that the underlying fitness landscape is stable along the phylogeny, taking into account only non-adaptive evolution, thus a mutation with a positive $\Sphy$ would be driving the current site toward are fitter amino-acid, and such mutation are effectively considered under maintenance positive selection.
        }
        \label{fig:method}
    \end{figure*}

    \subsection*{SNPs which are predicted to be advantageous are positively selected at the population scale}

    For each non-synonymous SNP observed at the population level, we predicted its scaled selection coefficient ($\Sphy$) from amino acid fitnesses estimated at the mammalian scale (fig.~\ref{fig:method}B).
    Then we ask whether the variants that are predicted to be advantageous, thus supposedly under maintenance selection, are showing signatures of short term positive selection.
    For example, if a variant is found to be fixed in many mammals but not in ancestral humans, and a mutation in humans is toward the mammalian version, is this mutation advantageous for the individual carrying it?
    At the population level (e.g.~in humans of Asian descent in fig.~\ref{fig:homo-afr-results}), we divided all observed SNPs in 5 classes of predicted selection coefficient: severely deleterious with $\divStrongDel$, deleterious with $\divDel$, weakly deleterious with $\divWeakDel$, weakly advantageous with $\divWeakAdv$ and advantageous with $\divAdv$ (fig.~\ref{fig:homo-afr-results}B).
    In each of theses classes, SNPs are then grouped by the number of derived alleles in the population (fig.~\ref{fig:homo-afr-results}C), generating a so called site-frequency spectrum (SFS).
    Compared to synonymous mutations which are supposedly neutral (black line in fig.~\ref{fig:homo-afr-results}C), mutations predicted to be deleterious (blue and green lines in fig.~\ref{fig:homo-afr-results}C) are underrepresented at high frequencies in the population, a signature of ongoing purifying selection.
    However, these observations remain qualitative.

    To obtain a quantitative estimate of the distribution of selection coefficient for each category of SNPs, we used the model polyDFE\cite{tataru_inference_2017, tataru_polydfe_2020}.
    This model uses the count of derived alleles for each frequency class of the SFS to infer the distribution of fitness effects (DFE) of the region under study.
    The model uses both the number of sites ($L$) for each category (fig.~\ref{fig:homo-afr-results}A, see Material and Methods), and the counts of synonymous derived allele as a neutral model, which allow to control for demographic effects, and polarization errors when inferring the ancestral state.
    These population-based models cannot give a specific estimate of the selection coefficient at the population-genetic scale ($\Spop$) for each SNP, but by gathering information across many SNPs, these models can estimate the proportion of advantageous mutations ($\PpolyAdv$), of nearly-neutral mutations ($\PpolyNeutral$) and of deleterious mutations ($\PpolyDel$) (fig.~\ref{fig:homo-afr-results}D).
    We found that SNPs predicted to be deleterious given their selection coefficient $\Sphy$ across mammals are effectively purged at the population-genetic scale, demonstrating that purifying selection is largely predictable and amino acids with low fitness across mammals are effectively purged away in each population.
    Notably, SNPs predicted to be advantageous given their selection coefficient $\Sphy$, thus supposedly under maintenance selection, are indeed under positive selection at the population-genetic scale, confirming that there is a local ongoing selection towards amino acids that are restoring the optimally functioning protein.
    In between these two extremes, mutations predicted to be weakly selected (cyan and orange in fig.~\ref{fig:homo-afr-results}) are effectively composed of neutral, positively and negatively selected mutations.

    We replicated such experiment across 28 populations (fig.~\ref{fig:heatmap}), and found that mutations predicted to be deleterious are indeed purged at the population-genetic scale (blue and green rows in fig.~\ref{fig:heatmap}A).
    However, mutations predicted to be weakly advantageous ($\divWeakAdv$) are composed of a mix of neutral and selected mutations with varying proportions across the different populations (orange row in fig.~\ref{fig:heatmap}B).
    Similarly, mutations predicted to be advantageous ($\divAdv$, under maintenance selection) are found to be positively selected with varying proportions across populations (red row in fig.~\ref{fig:heatmap}C).
    In other words, mutations predicted to be deleterious by the mutation-selection model are effectively purified away in any population, while mutations predicted to be advantageous can be either neutral or advantageous depending on the population.
    These variable proportions between populations can be explained by the effective number of individuals in the population ($\Ne$), a major driver of selection efficacy.
    Since $\Ne$ is not directly measurable, a reasonable proxy usually used is the within a population (Watterson $\theta_w$).
    We thus ordered populations by their levels of synonymous diversity (fig.~\ref{fig:diversity}A) and confirmed that higher synonymous diversity is typically accompanied by a smaller proportion of neutral mutations.
    This result suggests that populations with higher diversity (e.g.~\textit{Bos} or \textit{Ovis}) are more likely to discriminate between advantageous and deleterious mutations (fig~S1-S5).
    Alternatively stated, mutations in populations with low diversity (e.g.~\textit{Homo}) are effectively nearly-neutral and behave as would a neutral mutation.
    This result is qualitatively in accordance with the nearly-neutral theory of evolution which argues that mutations are less efficiently selected for in small populations.
    %, especially for mutations already predicted to be nearly-neutral at the phylogenetic scale (fig~S8-S9).

>>>>

\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\textrm{Var}}
\newcommand{\der}{\textrm{d}}
\newcommand{\e}{\textrm{e}}
\newcommand{\avg}[1]{\left< #1 \right>} % for average
\newcommand{\Ne}{N_{\textrm{e}}}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\textrm{fix}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\Sphy}{S_{0}}
\newcommand{\Sphyclass}{\mathcal{C}}
\newcommand{\SphyMean}{\overline{\Sphy}}
\newcommand{\divStrongDel}{\Sphy < -3}
\newcommand{\divDel}{-3 < \Sphy < -1}
\newcommand{\divWeakDel}{-1 < \Sphy < 0}
\newcommand{\divWeakAdv}{0 < \Sphy < 1}
\newcommand{\divAdv}{ \Sphy > 1}
\newcommand{\PdivStrongDel}{\proba \left[ \divStrongDel \right]}
\newcommand{\PdivDel}{\proba \left[ \divDel \right]}
\newcommand{\PdivWeakDel}{\proba \left[ \divWeakDel \right]}
\newcommand{\PdivWeakAdv}{\proba \left[ \divWeakAdv \right]}
\newcommand{\PdivAdv}{\proba \left[ \divAdv \right]}
\newcommand{\given}{\mid}
\newcommand{\Spop}{S}
\newcommand{\SpopMean}{\overline{\Spop}}
\newcommand{\polyDel}{\Spop < -1}
\newcommand{\polyNeutral}{-1 < \Spop < 1}
\newcommand{\polyAdv}{ \Spop > 1}
\newcommand{\PpolyDel}{\proba \left[ \polyDel \right]}
\newcommand{\PpolyNeutral}{\proba \left[ \polyNeutral \right]}
\newcommand{\PpolyAdv}{\proba \left[ \polyAdv \right]}

@@

\title{Up to 25\% of beneficial mutations in protein sequences are not adaptive innovations in mammals}

@@

    \begin{abstract}
        Mutations can be beneficial by bringing a new innovation to their bearer, allowing them to adapt to change in environments or in selection.
        However, mutations can also be beneficial because they are repairing previous deleterious changes, simply restoring existing functions
        In this study, we first estimated selective effects of mutations inside protein coding sequence, under a model assuming no adaptation at the phylogenetic scale.
        We then estimated the proportion of beneficial mutations that are not adaptive innovations, and subsequently estimated their proportion among all beneficial mutations at the population scale.
        Our work confirms that deleterious substitutions have accumulated in mammals and are currently being eliminated, resulting in up to 25\% of beneficial mutations that are not adaptive innovations, but instead are repairing previous deleterious changes.

    \end{abstract}

    \keywords{Adaptation \and beneficial mutations \and phylogenetic \and population-genetics \and codon models }

    Adaptation is considered as the main process driving the diversity of forms and functions across the tree of life, as it enables species to access new ecological niches and also respond to changes in their environment\cite{darwin_origin_1859}.
    In this context of adaptative evolution, a change in environment is echoed as a change of species' forms and functions, ultimately leaving signatures of accelerated changes in the species' genome\cite{merrell_adaptive_1994}.
    This signature of an accelerated rate of changes, generating an excess of beneficial mutations is used to quantify adaptation (fig.~\ref{fig:fitness-landscape}A), for specific genes along a specific lineage\cite{mcdonald_adaptative_1991, smith_adaptive_2002a, welch_estimating_2006}.
    The current availability of large scale genomic data and the development of theoretical models allowed to detect and quantify adaption across genes and lineages, which is now a common practice in evolutionary biology\cite{yang_statistical_2000, eyre-walker_genomic_2006, moutinho_variation_2019}.
    Although these approaches led to a better understanding of the processes shaping molecular evolution, it had the effect to merely associate beneficial mutations with adaptive evolution, which should be dissociated since adaptive evolution is not the only process which can lead to beneficial mutations\cite{charlesworth_other_2007, mustonen_fitness_2009}.
    Indeed, even in a constant environment for which the population is already adapted, a deleterious mutation in an individual can reach fixation in the population by genetic drift\cite{Ohta1992}.
    Subsequently, a mutation restoring the ancestral state will thus be beneficial (fig.~\ref{fig:fitness-landscape}B), even though no change in the environment occurred\cite{hartl_compensatory_1996, sella_application_2005, mustonen_fitness_2009, cvijovic_fate_2015}.
    More precisely, this restoration of the ancestral state can happen at another loci, in which case it is referred as compensating mutations\cite{hartl_compensatory_1996, mustonen_fitness_2009}.
    Conversely, the restoration of the ancestral state can happen at the same locus, in which case it is referred as a beneficial back-mutation\cite{piganeau_estimating_2003, charlesworth_other_2007}, which are the focus of this study.
    Taken together, not all beneficial mutations contribute to genetic innovation, and truly adaptive evolution can only be well estimated under the background of beneficial back-mutations\cite{keightley_what_2010, rice_evolutionarily_2015}.
    It is thus essential to distinguish the underlying reason for the advantage brought by a beneficial mutation, a back-mutation repairing damaged genomes or creating an adaptive innovations.

    For an individual carrying a beneficial mutation, whether this mutation is new innovation as response to a change in environment or a beneficial back-mutation is indistinguishable.
    Likewise, at the level of the population, both will result in a positive transmission bias of the beneficial allele.
    However, their consequences at the macro-evolutionary scale are fundamentally different.
    While adaptive evolution promotes phenotype diversification (fig.~\ref{fig:fitness-landscape}C), beneficial back-mutations promotes phenotype stability and preserves well established biological systems (fig.~\ref{fig:fitness-landscape}D).
    Fundamentally, adaptive evolution is unpredictable because it caused by an unpredictable change in the environment and the underlying fitness landscape\cite{bazykin_changing_2015}.
    On the other hand, beneficial back-mutations are predictable if the underlying stable fitness landscape is known.
    As a result, a way to quantify the amount of back-mutation is to estimate the underlying fitness landscape, assuming it is constant and stable across time.

    In this study, fitness landscapes of protein-coding genes are estimated from multi-species sequence alignments.
    Multi-species sequence alignments can be used as a proxy for fitness, since for example a deleterious amino acid will be more likely to be absent in the alignment.
    One of the most used methods to make these predictions is SIFT (Sort Intolerant From Tolerant), which reliably detects deleterious mutations in many taxa\cite{ng_sift_2003, vaser_sift_2016}.
    However, SIFT scores is not interpretable as a selection coefficient.
    Second, we know that a lot of mechanisms can influence amino acid frequency other than fitness (phylogenetic inertia, mutation biases, GC-biased gene conversion, mutational paths) which are not accounted for in SIFT\@.
    Contrarily, so-called mutation-selection codon models are rooted within a population-genetic formalism\cite{halpern_evolutionary_1998, mccandlish_modeling_2014}.
    These phylogenetic codon models can disentangle processes of mutation at the DNA level and selection at the amino acid level.
    Such codon models provide a nearly-neutral framework of evolution by estimating the fitness landscape over amino acid sequences for each site of the sequence (fig~\ref{fig:method}A) while taking into account non-adaptive processes impacting codon frequency\cite{halpern_evolutionary_1998, rodrigue_mechanistic_2010, tamuri_estimating_2012, latrille_improved_2022a}.
    From the fitness landscape obtained by fitting the model to the data, one can thus compute the scaled selection coefficient, denoted $\Sphy$, of any non-synonymous mutations.
    Importantly, mutation-selection codon models assume that the underlying fitness landscape is stable along the phylogeny (as in fig.~\ref{fig:fitness-landscape}B), hence the null subscript $0$ for $\Sphy$.
    As a result, a mutation with a positive $\Sphy$ would be driving the current site toward a fitter amino-acid, and thus considered a beneficial back-mutation repairing existing functions.
    Altogether, our expectation if that beneficial back-mutations are detectable and scattered across the genomes, while at the same time some slightly deleterious mutations are reaching fixations at other loci, generating a balance at which genomes are constantly both damaged and repaired at different loci.
    Even though each of these back-mutations is expected to have a small beneficial effect on their bearer, and that our assumptions may not always be valid, we expect that the genome-wide signature of beneficial back-mutations can be detected and quantified.

    By integrating datasets at different scales (phylogeny and population), we can predict whether substitutions in terminal lineages or any mutation currently segregating in a population are beneficial back-mutations, i.e with a positive $\Sphy$ (fig~\ref{fig:method}B).
    Such a dataset can allow us to address different questions regarding the selection affecting protein-coding DNA sequences.
    Are we empirically observing beneficial back-mutations that reached fixation recently, and one that are currently segregating in populations?
    Are they effectively beneficial for the individual carrying it?
    Conversely, what is the probability for a beneficial mutation to be a back-mutation?
    In other words, can we find empirical evidence and can we quantify the amount of beneficial mutations that are restoring damaged genomes instead of creating adaptive innovations.
    To answer these questions, we estimated the fitness landscape of amino-acids for long-term evolution using available multiple sequence alignments between mammalian species\cite{ranwez_orthomam_2007, howe_ensembl_2021}.
    We then integrated data at the phylogenetic and population scale to estimate the amount of beneficial mutations across the entire exome for 28 populations from 6 genera (\textit{Equus},  \textit{Bos}, \textit{Capra}, \textit{Ovis}, \textit{Chlorocebus} and \textit{Homo}; fig.~\ref{fig:method}A\&B).

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.fitness-landscape}
        \caption{
            Panel A and B: at a given site inside a protein-coding DNA sequence, the different amino acids (x-axis) have different fitnesses (y-axis).
            Under a fluctuating fitness landscape (panel A), these fitnesses are changing with time.
            The protein sequence is always lagging behind the moving target defined by the amino acid fitnesses, and since substitutions are accepted preferentially if they are in the direction of this target, substitutions are on average adaptive.
            At the phylogenetic scale (panel C), this phenomena promotes phenotype diversification across species.
            Under a stable fitness landscape (panel B), all the mutations reaching fixation are either slightly deleterious and reaching fixation due to drift, or are beneficial back-mutations restoring a more optimal amino acid.
            Scarce beneficial back-mutations are likely to reach fixation and a bulk of deleterious mutations are eventually reaching fixation due to their sheer mass.
            Across the genome, this process generate a balance at which genomes are constantly both damaged and repaired at different loci.
            At the phylogenetic scale (panel D), this phenomena promotes promotes phenotype stability and preserves well established biological systems.
        }
        \label{fig:fitness-landscape}
    \end{figure*}

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.method.proba}
        \caption{
            Integrating datasets at different scales.
            At the phylogenetic level (panel A), the amino acid Wrightian fitness for each site is estimated from protein-coding DNA alignments using mutation-selection codon models.
            Mutation-selection models assume a stable fitness landscape, taking into account only non-adaptive evolution.
            A mutation with a $\Sphy > 1$ would be driving the current site toward are fitter amino-acid, and such mutation are considered a beneficial back-mutation repairing existing functions (panel B).
            At the population scale (panel C), each single nucleotide polymorphism (SNP) segregating in the population can be classified as a beneficial back-mutation (in red) or not.
            The number of derived allele allow to estimate the probability of beneficial mutations at the population scale, for all SNPs or among back-mutations (panel D).
            The probability of back-mutation is obtained as their fraction among all possible mutations from the ancestral genome sequence (panel E).
            Altogether, we can estimate the probability of back-mutations among all beneficial ones, using Bayes' theorem (panel F).
        }
        \label{fig:method}
    \end{figure*}

    \subsection*{Back-mutations in the terminal lineages}
    We first probed for empirical evidence of beneficial back-mutation in the terminal lineages, which should be effectively beneficial.
    We thus reconstructed the substitutions occurring along the terminal branches of the phylogenetic tree.
    For each substitution, we estimated its scaled selection coefficient ($\Sphy$) based on the stable fitness landscape of amino-acids estimated at the scale of the mammals' phylogeny (fig.~\ref{fig:method}A).
    We showed that among all the substitutions in the terminal lineages, between 10 and 13\% had $\Sphy > 1$ (table SX) and between 25 and 30\% had $\Sphy > 0$ (table SX).
    Substitutions with $\Sphy > 1$ have been beneficial since the ratio of non-synonymous over synonymous divergence, called $\dnds$, was estimated to be between 1.15 and 1.68 in the different lineages (table SX).
    This meant that beneficial back-mutations with $\Sphy > 1$ reach fixation more frequently than synonymous mutations, which are supposed to be neutral, thus are effectively beneficial.
    This also means that the ratio $\dnds$ usually used to detect adaptation while relying on substitutions between close species\cite{mcdonald_adaptative_1991, galtier_adaptive_2016} or at the phylogenetic scale\cite{goldman_codonbased_1994, yang_codonsubstitution_2002} is overinflated due to beneficial back-mutations.
    Then, $\dnds$ computed exclusively on non-synonymous sites with a $\Sphy < 0$ only represent the rate of evolution discarding beneficial back-mutations by construction.
    Thus, if we divide the $\dnds$ estimated on all substitutions by the one estimated only on substitutions predicted to be deleterious, we directly get the proportion of the rate of evolution which is contributed by beneficial back-mutations (see methods).
    This way, we estimated that between 22\% and 27\% of $\dnds$ is inflated because of beneficial back-mutations (table SX).

    \subsection*{Back-mutations in the populations}
    Given that beneficial back-mutations are effectively selected for, we then ask which proportion of new mutations are beneficial back-mutations.
    For each population with available polymorphism data, we first reconstructed the ancestral sequence (instead of the reference sequence) at the base of the population genealogy.
    From this reference sequence, we computed the scaled selection coefficient ($\Sphy$) for each possible non-synonymous mutations.
    We thus obtained the distribution of fitness effects (DFE) of new mutations from the ancestral sequence in \textit{Homo sapiens} (fig.~\ref{fig:homo-afr-results}A), under the assumption of the stable fitness landscape.
    Because $\Sphy$ is based on estimation of fitnesses at the mammalian scale, mutations with $\Sphy>1$ are considered beneficial back-mutations.
    Beneficial back-mutations only represents ~1.5\% of all possible new mutations (red in fig.~\ref{fig:homo-afr-results}A, table~SX).
    However, when focusing on the mutations that are observed in populations, namely single nucleotide polymorphism (SNP), beneficial back-mutations represents ~8.5\% of all observed mutations (red in fig.~\ref{fig:homo-afr-results}B, table~SX).

    Additionally, in populations, the frequencies at which SNPs are segregating is informative of their selective effect, e.g. deleterious SNPs are segregating at lower frequencies because they are purified.
    As such, methods have been developed to estimate the distribution of fitness effect (DFE) at the population scale by gathering information across many SNPs\cite{eyre-walker_distribution_2006a, eyre-walker_estimating_2009a, galtier_adaptive_2016, tataru_inference_2017}.
    These methods offer a unique opportunity to compare the DFE at the population scale ($\Spop$) and at the phylogenetic scale ($\Sphy$).
    In order to estimate the DFE at the population level and gather robust statistical estimates, we grouped SNPs observed at the population level (e.g.~in humans of African descent in fig.~\ref{fig:homo-afr-results}) in 5 classes of predicted selection coefficient: severely deleterious with $\divStrongDel$, deleterious with $\divDel$, weakly deleterious with $\divWeakDel$, weakly beneficial with $\divWeakAdv$ and beneficial with $\divAdv$ (fig.~\ref{fig:homo-afr-results}B).
    In each of theses classes, the number of derived alleles in the population for each SNPs is used to generate a site-frequency spectrum (SFS) (fig.~\ref{fig:homo-afr-results}C).
    Compared to synonymous mutations which are supposedly neutral (black line in fig.~\ref{fig:homo-afr-results}C), mutations predicted to be deleterious (blue and green lines in fig.~\ref{fig:homo-afr-results}C) were, as expected, underrepresented at high frequencies in the population, a signature of ongoing purifying selection.
    However, these observations remain qualitative.

    To obtain a quantitative estimate of the distribution of selection coefficient for each category of SNPs, we used the model polyDFE\cite{tataru_inference_2017, tataru_polydfe_2020}.
    This model uses the count of derived alleles for each frequency class of the SFS to infer the distribution of fitness effects (DFE) of the region under study.
    The model uses the number of sites ($L$) for each category (fig.~\ref{fig:homo-afr-results}A, see Material and Methods), the counts of synonymous derived allele as a neutral model, which allow to control for demographic effects, and polarization errors when inferring the ancestral state.
    We could estimate the proportion of beneficial mutations ($\PpolyAdv$), nearly-neutral mutations ($\PpolyNeutral$) and deleterious mutations ($\PpolyDel$) for each class of selection previously defined (fig.~\ref{fig:homo-afr-results}D).
    We found that SNPs predicted to be deleterious based on the selection coefficient $\Sphy$ estimated across the phylogeny of mammals were effectively purged at the population scale.
    Purifying selection is therefore largely predictable and amino acids with low fitness across mammals are effectively purged away in each population.
    Then, SNPs predicted to be beneficial based on the selection coefficient $\Sphy$ estimated at the scale of the phylogeny of mammals are indeed beneficial for individuals currently bearing them.
    These results confirmed that selection towards amino acids that are restoring existing function is ongoing in these populations.
    Between these two extreme cases, SNPs predicted to be weakly selected (cyan and orange in fig.~\ref{fig:homo-afr-results}) were effectively composed of neutral, beneficial and deleterious mutations.

    We replicated our experiment across 28 populations of different mammal species (fig.~\ref{fig:heatmap}) and found that SNPs predicted to be deleterious were indeed purged at the population scale (blue and green rows in fig.~\ref{fig:heatmap}A).
    However, SNPs predicted to be weakly beneficial ($\divWeakAdv$) were composed of a mix of neutral and selected mutations with varying proportions across the different populations (orange row in fig.~\ref{fig:heatmap}B).
    Finally, SNPs predicted to be beneficial back-mutations ($\divAdv$) were found to be beneficial with varying proportions across populations (red row in fig.~\ref{fig:heatmap}C).
    The variable proportions found between populations can be explained by the effective number of individuals in the population ($\Ne$), which is well-known to be a major driver of selection efficacy.
    Since $\Ne$ is not directly measurable, we used the Watterson $\theta_w$ as a proxy to order populations by their levels of synonymous diversity.
    Higher synonymous diversity is associated with a smaller proportion of neutral mutations (fig.~\ref{fig:diversity}A).
    This result confirms the nearly-neutral theory and suggests that populations with higher diversity (e.g.~\textit{Bos} or \textit{Ovis}) are more likely to discriminate between beneficial and deleterious mutations (fig~SX-SX), while mutations in populations with low diversity (e.g.~\textit{Homo}) are effectively nearly-neutral.

    \subsection*{Proportion of back-mutations among all beneficial mutations}
